---
title: "geomx_analysis"
output: rmarkdown::github_document
date: '2022-06-29'
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, root.dir = "C:/Users/Strix/GitRepos/NOD_spt_pd1")
knitr::opts_knit$set(echo = TRUE, root.dir = "C:/Users/Strix/GitRepos/NOD_spt_pd1")
```

## GeoMX analysis

In this analysis, we looked at various proteins within beta islets of NOD mice age 16-22 weeks of age. The three treatment groups examined include healthy **isotype antibody-treated (IgG)**; **spontaneously diabetic (Spt)**; and **PD-1 blockade-treated diabetic (PD1)** mice. Both IgG and PD1 mice were treated with two doses of 200ug of antibody (anti-PD-1 clone 29F.1A12 or isotype IgG2a control) on days 0,2 before sacrifice on day 4. Pancreata were excised and immediately preserved in 10% formalin for 24h, stored in 70% ethanol, embedded in paraffin, and sectioned for H&E staining or staining with anti-INS or anti-CD45 antibodies. Both CD45 and INS staining was used to differentiate various regions within the islets. Islets were scored as healthy with no immune infiltrate (score 0), per-insulitis (score 1-2), or overt insulitis (score 3-4).

```{r libraries, message = FALSE, warning = FALSE}
require(dplyr, quietly = T) #simple manipulation of data.frames
require(reshape, quietly = T) #melting and casting data.frames
require(ggplot2, quietly = T) #plotting
require(readxl, quietly = T) #reading excel spreadsheets
require(rstatix, quietly = T) #to add significance to p values
require(ggprism, quietly = T) #adding p values to plots
require(patchwork, quietly = T) #adding p values to plots
require(magrittr, quietly = T) #adding p values to plots
require(gridExtra, quietly = T) #display multiple plots
require(vegan , quietly = T) #principal component analysis
require(nlme, quietly = T) #mixed linear model analysis
require(ggrepel, quietly = T) #labels for pca plots
require(tibble, quietly = T) #rownames_to_column function
require(useful, quietly = T) #corner function to peek at data.frames
```

```{r define color palettes}
treatment_colors <- c("IgG" = "black", 
                      "Spt" = rgb(253,128,8, maxColorValue = 255), 
                      "PD1" = rgb(128,0,128, maxColorValue = 255))
islet_score_colors <- c("Insulitis" = "darkred",
                  "Peri.insulitis" = "royalblue3",
                  "Normal" = "seagreen4")
```

Read in the data files. One data file contains the code-friendly protein names, description, and associated mouse genes. Two data files for CD45+ and INS+ regions of interest describing the normalized (using the GeoMX software) quantification of protein expression. The data for the CD45 and ins regions are merged, as they are already distinguished based on staining for CD45 and INS.

```{r read in data files}
#read in CD45 data that has been normalized
cd45 <- as_tibble(t(read_excel(file.path("data/geoMX/norm_geoMean_CD45_trimmed.xlsx"))))
colnames(cd45) <- cd45[1,] #set column names manually (doesn't work in read_excel)
cd45 <- cd45[-1,] #remove row with column names
# cd45$islet_score <- gsub("\\-", ".", cd45$islet_score) #change dashes in Peri-insulitis to periods (Peri.insulitis) in data if not already
# colnames(cd45) <- gsub("\\-", ".", colnames(cd45)) #change dashes in protein colnames in data if not already

#read in CD45 data that has been normalized
ins <- as_tibble(t(read_excel(file.path("data/geoMX/norm_geoMean_INS_trimmed.xlsx"))))
colnames(ins) <- ins[1,] #set column names manually (doesn't work in read_excel)
ins <- ins[-1,] #remove row with column names
#ins$islet_score <- gsub("\\-", ".", ins$islet_score) #change dashes in Peri-insulitis to periods (Peri.insulitis) in data if not already
#colnames(ins) <- gsub("\\-", ".", colnames(ins)) #change dashes in protein colnames in data if not already

#merge the CD45 and INS data
geomx <- rbind(cd45,ins)
colnames(geomx)
geomx <- geomx[,c("scan_id","ROI",colnames(geomx)[26:99])] #remove excess columns
corner(geomx)

#read in gene annotations file that includes additional information for proteins detected
gene_annot <- read_excel(file.path("data/geoMX/gene_annotations.xlsx"), col_names = T)
#gene_annot$protein <- gsub("\\-", ".", gene_annot$protein) #change dashes in protein colnames in data if not already
gene_annot$target_group <- gsub("All Targets;","",gene_annot$target_group) #remove extraneous text from description
head(gene_annot)
```

Tally the number of samples to get an idea of what the data looks like. Note that there is only one normal islet in the CD45 group that was analyzed, so that will be removed before any statistical analyses as it doesn't make sense - normal islets should have no CD45+ infiltrate. Note that there aren't many samples for the PD1 treatment group.
```{r exploration of sample numbers}
geomx %>% group_by(treatment,islet_score,label) %>% tally() %>% cast(label+islet_score ~ treatment, value = "n", fill = 0)
geomx <- geomx %>% filter(islet_score != "Normal" & label == "CD45" | label == "INS")
```

Examine protein expression broadly across all of the samples based on their region of classification (CD45+ or INS+ region) to confirm that expression levels generally make sense. CD45+ regions should have higher expression of immune-cell associated proteins (CD3e, CD45, CD8a, PD-1, etc.) compared to INS+ regions. Multiple t tests are used with a Benjamini & Hochberg "BH" correction.

```{r differences in protein expression by region classification (CD45/INS), out.width="900px"}
#melt data to long format for plotting
colnames(geomx)
melted <- data.frame(geomx) %>% melt(id.vars = c("treatment","mouse","islet_score","label","scan_id","ROI"))
colnames(melted)[which(colnames(melted) == "variable")] <- "protein"
colnames(melted)[which(colnames(melted) == "value")] <- "expression"
head(melted)

#correct the class of some of the variables
melted$expression <- as.numeric(melted$expression)
melted$protein <- factor(melted$protein, levels = melted %>% 
                           group_by(protein) %>% 
                           summarize(avg = mean(expression)) %>% 
                           arrange(desc(avg)) %>% 
                           pull(var = protein))
melted$treatment <- factor(melted$treatment, levels = c("IgG","Spt","PD1"))
melted$islet_score <- factor(melted$islet_score)

melted <- merge(melted, gene_annot[,c("protein","gene","code_class")], by = c("protein")) #add protein annotation info

#calculate the p values for plotting using t tests
p_vals <- melted %>%
  rstatix::group_by(protein) %>%
  rstatix::t_test(expression ~ label) %>%
  rstatix::adjust_pvalue(p.col = "p", method = "BH") %>%
  rstatix::add_significance(p.col = "p.adj") %>% 
  rstatix::add_xy_position(x = "protein", dodge = 0.8) # important for positioning!
p_vals$y.position <- log2(p_vals$y.position) #log2 transformed data will be plotted, so djust pvalue position

#plot the expression data as violin plots grouped by region classification
melted %>% 
  ggplot(aes(x = protein, 
             y = log2(expression), #log2 expression values are used
             color = label, 
             fill = label)) + 
  geom_violin(alpha = 0.3, 
              adjust = 2, 
              scale = "width", 
              draw_quantiles = c(0.5), #draw the median as a line
              position = position_dodge(0.8)) + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.x = element_blank()) +
  labs(title = "GeoMX protein expression: CD45+/INS+ regions") +
  add_pvalue(p_vals %>% filter(p.adj < 0.05),  #only plot significant values
             color = "black",
               xmin = "xmin", 
               xmax = "xmax",
               label = "p.adj.signif",
               tip.length = 0,
             show.legend = F,
             inherit.aes = F)
```

Expression of CD45 and other lymphocyte-associated proteins (CD19, CD11c, F4/80, CD3e, etc) are higher in the CD45 regions as expected. Look at differences between islets with peri-insulitis and overt insulitis. Multiple t tests are used with a Benjamini & Hochberg "BH" correction.
```{r differences in protein expression by islet score (Insulitis/Peri.Insulitis), out.width="900px"}
p1_vals <- melted %>%
  filter(label == "CD45") %>%
  rstatix::group_by(protein) %>%
  rstatix::t_test(expression ~ islet_score) %>%
  rstatix::adjust_pvalue(p.col = "p", method = "BH") %>%
  rstatix::add_significance(p.col = "p.adj") %>% 
  rstatix::add_xy_position(x = "protein", dodge = 0.8) #important for positioning!
p1_vals$y.position <- log2(p1_vals$y.position) #log2 transformed expression values are used, so this adjustment is needed
p1_vals <- merge(p1_vals, gene_annot[,c("protein","code_class")], by = "protein")

p1 <- melted %>%
    filter(label == "CD45") %>%
    ggplot(aes(x = protein, 
               y = log2(expression), 
               color = islet_score, 
               fill = islet_score)) +
    geom_violin(alpha=0.3, 
                adjust=2, 
                scale = "width", 
                draw_quantiles = c(0.5), 
                position = position_dodge(0.8)) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, 
                                     vjust = 0.5, 
                                     hjust=1),
          axis.title.x = element_blank()) +
    scale_color_manual(values = islet_score_colors) +
    scale_fill_manual(values = islet_score_colors) +
    { if (sum(p1_vals$p.adj < 0.05) > 0) #check if there are any significant p values, otherwise, don't plot any NS
      add_pvalue(p1_vals %>% filter(p.adj < 0.05),  #only plot significant values
               color = "black",
               xmin = "xmin",
               xmax = "xmax",
               label = "p.adj.signif",
               tip.length = 0,
               show.legend = F,
               inherit.aes = F) } +
    labs(title = "GeoMX protein expression (CD45+ regions)")


p2_vals <- melted %>%
  filter(label == "INS") %>%
  rstatix::group_by(protein) %>%
  rstatix::t_test(expression ~ islet_score) %>%
  rstatix::adjust_pvalue(p.col = "p", method = "BH") %>%
  rstatix::add_significance(p.col = "p.adj") %>% 
  rstatix::add_xy_position(x = "protein", dodge = 0.8) # important for positioning!
p2_vals$y.position <- log2(p2_vals$y.position) #log2 transformed expression values are used, so this adjustment is needed
p2_vals <- merge(p2_vals, gene_annot[,c("protein","code_class")], by = "protein")

p2_vals %>% filter(p.adj < 0.05) %>% arrange(p.adj) %>% head(15) #show some of the top DE proteins

p2 <- melted %>%
    filter(label == "INS") %>%
    ggplot(aes(x = protein, 
               y = log2(expression), 
               color = islet_score, 
               fill = islet_score)) +
    geom_violin(alpha=0.3, 
                adjust=2, 
                scale = "width", 
                draw_quantiles = c(0.5), 
                position = position_dodge(0.8)) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, 
                                     vjust = 0.5, 
                                     hjust=1),
          axis.title.x = element_blank()) +
    scale_color_manual(values = islet_score_colors) +
    scale_fill_manual(values = islet_score_colors) +
    { if (sum(p2_vals$p.adj < 0.05) > 0) #check if there are any significant p values, otherwise, don't plot any NS
      add_pvalue(p2_vals %>% filter(p.adj < 0.05), #only plot significant values
               color = "black",
               xmin = "xmin",
               xmax = "xmax",
               label = "p.adj.signif",
               tip.length = 0,
               show.legend = F,
               inherit.aes = F) } +
    labs(title = "GeoMX protein expression (INS+ regions)")

grid.arrange(p1, p2, nrow=2)
```

There are no significant differences between CD45+ regions within islets with insulitis vs. peri-insulitis. The only identified differences were between INS+ normal islets (in IgG mice) and those with peri-insulitis or insulitis. Most DE proteins in the INS+ regions are immune cell-associated (CD40, CD45, VISTA, CD11c) which indicates that there is some contamination of these regions with immune infiltrate. However, the immune cells in this region are likely to be associated with the "leading edge" of the infiltrate that are closest to the INS+ cells.

```{r PCA analysis CD45 regions}
df_for_pca <- geomx %>% filter(label == "CD45")
df_for_pca[,-c(1:6)] <- sapply(df_for_pca[,-c(1:6)], as.numeric)
df_for_pca[,-c(1:6)] <- sapply(df_for_pca[,-c(1:6)], log2)

prin_comp <- rda(df_for_pca[,-c(1:6)])
pca_scores <- scores(prin_comp)

# Define shapes for plot
shapes = c(16, 17, 15) 
shapes <- shapes[as.numeric(factor(df_for_pca$islet_score))]

#plot using base R
plot(pca_scores$sites[,1],
     pca_scores$sites[,2],
     pch = shapes,
     col = paste(treatment_colors[df_for_pca$treatment]),
     xlim = c(-3,3),
     ylim = c(-3,3),
     cex=3)
ordiellipse(prin_comp, factor(df_for_pca$treatment, levels = c("IgG","Spt","PD1")), 
            conf=0.90, 
            col = treatment_colors, 
            lwd = 2)

#keep data to later plot in ggplot
df_all <- cbind(df_for_pca, pca_scores$sites)
pca_weights <- data.frame(pca_scores$species, label = "CD45") %>% 
                          rownames_to_column("protein")

#collect variance explained for each PC for labeling
variance_explained_labels <- c(paste0("PC1 (", 
                                      round(summary(prin_comp)$cont$importance[2, "PC1"]*100, 2),
                                      "% of variance)"),
                               paste0("PC2 (", 
                                      round(summary(prin_comp)$cont$importance[2, "PC2"]*100, 2),
                                      "% of variance)"))
```

```{r PCA analysis INS}
df_for_pca <- geomx %>% filter(label == "INS")
df_for_pca[,-c(1:6)] <- sapply(df_for_pca[,-c(1:6)], as.numeric)
df_for_pca[,-c(1:6)] <- sapply(df_for_pca[,-c(1:6)], log2)

prin_comp <- rda(df_for_pca[,-c(1:6)])
pca_scores <- scores(prin_comp)

# Define shapes for plot
shapes = c(16, 17, 15) 
shapes <- shapes[as.numeric(factor(df_for_pca$islet_score))]

plot(pca_scores$sites[,1],
     pca_scores$sites[,2],
     pch = shapes,
     col = paste(treatment_colors[df_for_pca$treatment]),
     xlim = c(-4,3),
     ylim = c(-2,4),
     cex=3)
ordiellipse(prin_comp, 
            factor(df_for_pca$treatment, levels = c("IgG","Spt","PD1")), 
            conf=0.90, 
            col = treatment_colors, 
            lwd = 2)

#gather data for PCA plot in ggplot
df_for_pca <- cbind(df_for_pca, pca_scores$sites)
df_all <- rbind(df_all, df_for_pca)
pca_weights <- rbind(pca_weights, 
                     data.frame(pca_scores$species, label = "INS") %>% 
                     rownames_to_column("protein"))

#collect variance explained for each PC (as a percentage of 100, rounded to two decimals) for labeling axes in plots
variance_explained_labels <- c(variance_explained_labels,
                               paste0("PC1 (", 
                                      round(summary(prin_comp)$cont$importance[2, "PC1"]*100, 2),
                                      "% of variance)"),
                               paste0("PC2 (", 
                                      round(summary(prin_comp)$cont$importance[2, "PC2"]*100, 2),
                                      "% of variance)"))
names(variance_explained_labels) <- c("PC1_CD45","PC2_CD45","PC1_INS","PC2_INS") #name the vector of axis titles
```

These plots can be generated using ggplot2 which looks nicer and allows for increased customization. Both the PCA plots will be generated separately to allow for the axes to be labeled with the customized variance labels (rather than facetting the plots).

```{r ggplot of PCA plots}
pca_cd45 <- df_all %>%
    filter(label == "CD45") %>%
    ggplot(aes(x = PC1, 
               y= PC2, 
               color = treatment, 
               shape = islet_score, 
               group = treatment)) +
    geom_point(size = 4, 
               alpha = 0.6) +
    stat_ellipse(level = 0.9) + #90% confidence intervals
    scale_color_manual(values = treatment_colors) +
    theme_minimal() +
    theme(panel.grid = element_blank(),
          axis.text = element_blank(),
          panel.border = element_rect(color = "black", fill = NA),
          text = element_text(size = 15),
          legend.position = "none") + #this plot will be on the left so it doesn't need a legend
    labs(title = "CD45+ regions of islets",
         x = variance_explained_labels["PC1_CD45"],
         y = variance_explained_labels["PC2_CD45"])
    
pca_ins <- df_all %>%
    filter(label == "INS") %>%
    ggplot(aes(x = PC1, y= PC2, 
               color = treatment, 
               shape = islet_score, 
               group = treatment)) +
    geom_point(size = 4, 
               alpha = 0.6) +
    stat_ellipse(level = 0.9) + #90% confidence intervals
    scale_color_manual(values = treatment_colors) +
    theme_minimal() +
    theme(panel.grid = element_blank(),
          axis.text = element_blank(),
          panel.border = element_rect(color = "black", fill = NA),
          text = element_text(size = 15)) +
    labs(shape = "Islet score", 
         color = "Treatment",
         title = "INS+ regions of islets",
         x = variance_explained_labels["PC1_INS"],
         y = variance_explained_labels["PC2_INS"])

grid.arrange(pca_cd45, pca_ins, nrow = 1, widths = c(1.5, 2))
```

```{r ggplot of PCA weights}
proteins_of_interest <- c("LAG3","Perforin","GZMB","CTLA4","Ki.67","ICOS","CD127","IFNGR","Tim3","BAD","GITR","PD.1","Icos","BCLXL","Cleaved_Caspase3","CD4","CD8a","VISTA")

pca_labels <- pca_weights %>% 
  filter(protein %in% proteins_of_interest)

weights_cd45 <- pca_weights %>%
    filter(label == "CD45") %>%
    ggplot(aes(x = PC1, y = PC2)) +
    facet_wrap(~ label, scales = "free") +
    scale_color_manual(values = treatment_colors) +
    theme_minimal() +
    theme(panel.grid = element_blank(),
          axis.text = element_blank(),
          panel.border = element_rect(color = "black", fill = NA),
          text = element_text(size = 15)) +
    geom_segment(aes(x = 0, y = 0, xend = PC1, yend = PC2),
                 arrow = arrow(length = unit(4, "mm")),
                 color = "black",
                 alpha = 0.5) +
    geom_text_repel(data = pca_labels %>% filter(label=="CD45"), aes(label = protein), 
                    color = "black",
                    max.overlaps = 4) +
    labs(x = variance_explained_labels["PC1_CD45"],
         y = variance_explained_labels["PC2_CD45"])

weights_ins <- pca_weights %>%
    filter(label == "INS") %>%
    ggplot(aes(x = PC1, y= PC2)) +
    facet_wrap(~ label, scales = "free") +
    scale_color_manual(values = treatment_colors) +
    theme_minimal() +
    theme(panel.grid = element_blank(),
          axis.text = element_blank(),
          panel.border = element_rect(color = "black", fill = NA),
          text = element_text(size = 15)) +
    geom_segment(aes(x = 0, y = 0, xend = PC1, yend = PC2),
                 arrow = arrow(length = unit(4, "mm")),
                 color = "black",
                 alpha = 0.5) +
    geom_text_repel(data = pca_labels %>% filter(label=="INS"), aes(label = protein), 
                    color = "black",
                    max.overlaps = 4) +
    labs(x = variance_explained_labels["PC1_INS"],
         y = variance_explained_labels["PC2_INS"])

grid.arrange(weights_cd45, weights_ins, nrow = 1)
```

Next, the differences between treatment groups will be analyzed by generating volcano plots. Here is an example of the comparison between PD1 and Spt. There are no significant differences.

```{r Protein expression between treatment groups}
meltednrml <- melted %>% filter(label == "CD45") %>% dplyr::select(-islet_score)
meltednrml$treatment <- relevel(meltednrml$treatment, ref = "PD1")
df <- data.frame()

for (current_protein in as.character(unique(meltednrml$protein))) {
  tmp <- meltednrml %>% 
    filter(protein == current_protein)
  tmp <- data.frame(summary(lme(expression ~ treatment, random = list(~1|mouse, ~1|scan_id), data = tmp))$tTable[-1,]) %>% 
    rownames_to_column("comparison") %>% 
    mutate(protein = current_protein, ref = "PD1", comparison = gsub("treatment", "", comparison))
  df <- rbind(df, tmp)
}

avgs <- meltednrml %>% 
  group_by(treatment,protein) %>% 
  summarize(avg = mean(expression), .groups = "keep")

tbl <- merge(df, avgs %>% cast(protein~treatment,value = "avg"), by = c("protein"))
tbl <- tbl %>% mutate(log2FC = ifelse(comparison == "Spt", log2(Spt/PD1), log2(IgG/PD1)))
pd1_v_spt <- tbl %>% filter(comparison == "Spt")
pd1_v_spt$p.adj <- p.adjust(pd1_v_spt$p.value, method = "BH")

pd1_v_spt %>%
    ggplot(aes(x=log2FC,y=-log10(p.adj))) + 
    theme_minimal() + 
    theme(panel.grid = element_blank(),
          panel.border = element_rect(color="grey10", fill = NA)) +
    geom_hline(yintercept = -log10(0.05), color = "grey90") + 
    geom_vline(xintercept = c(-0.5,0.5), color = "grey90") +
    geom_point() + 
    scale_x_continuous(limits = c(-3,3)) +
    scale_y_continuous(limits = c(0,2)) +
    geom_text_repel(data = subset(pd1_v_spt, p.adj < 0.5 & abs(log2FC) > 0.5), aes(label=protein))
```