---
title: "3_figure2_plots"
output: rmarkdown::github_document
date: '2022-07-10'
---

The purpose of this figure is to show that CD8+ T cells in the pancreas show a gain in proliferation, gain in exhaustion-like features, and loss of memory-like features in checkpoint-induced T1D compared to spontaneous T1D.

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, root.dir = "C:/Users/Strix/GitRepos/NOD_spt_pd1")
knitr::opts_knit$set(echo = TRUE, root.dir = "C:/Users/Strix/GitRepos/NOD_spt_pd1")
```

```{r libraries, message = FALSE, warning = FALSE}
require(dplyr, quietly = T) #simple manipulation of data.frames
require(Seurat, quietly = T) #scRNA-seq analysis
require(reshape, quietly = T) #melting and casting data.frames
require(ggplot2, quietly = T) #plotting
require(rstatix, quietly = T) #adding p values to plots
require(ggprism, quietly = T) #adding p values to plots
require(ggsci, quirtly = T) #IGV color palette
require(patchwork, quietly = T) #adding p values to plots
require(magrittr, quietly = T) #adding p values to plots
require(gridExtra, quietly = T) #display multiple plots
require(ggrepel, quietly = T) #labels for plots
require(tibble, quietly = T) #rownames_to_column function and deframe in GSEA analysis
require(presto, quietly = T) #wilcoxauc quicker test for GSEA analysis across treatment groups
require(msigdbr, quietly = T) #access to msigdb gene signatures
require(fgsea, quietly = T) #GSEA analysis
require(useful, quietly = T) #corner function to peek at data.frames
require(viridis, quietly = T) #viridis color palettes
require(readxl, quietly = T) #reading Excel files
require(msigdbr, quietly = T) #accessing gene signatures from mSigDB
source("scripts/utils.R") #custom color palettes and other functions
set.seed(1) #set a seed for reproducibility
```

```{r read in Seurats}
pan8 <- readRDS("objects/seurat_cd8_panonly_clustered.rds")
```

The purpose of these figures are to show that PD-1 blockade promotes CD8+ T cell clonal expansion, loss of a memory-like subset, and exhaustion-like transcriptional features in NOD pancreas.

```{r plot UMAP of different treatment groups}
pan8_cls <- pan8@meta.data %>% 
        dplyr::select(UMAP_1, UMAP_2, seurat_clusters) %>% 
        group_by(seurat_clusters) %>% 
        summarise(U1 = mean(UMAP_1), U2 = mean(UMAP_2)) %>% 
        data.frame()

p_cd8_panonly_cluster <- pan8@meta.data %>% 
                            ggplot(aes(x = UMAP_1, 
                                     y = UMAP_2, 
                                     color = seurat_clusters)) + 
                            geom_point(shape = 16, 
                                       alpha = 0.6) +
                            theme_minimal() +
                            theme(axis.title = element_blank(),
                                  axis.text = element_blank(),
                                  text = element_text(size = 15),
                                  panel.grid = element_blank()) +
                            labs(x = 'UMAP 1', 
                                 y = 'UMAP 2', 
                                 color = "Seurat clusters",
                                 title = "CD8+ T cells in the pancreas",
                                 size = "Size of clonotype\n(pancreas only)") +
                            scale_alpha_continuous(range = c(0.4,1)) +
                            scale_size_continuous(range=c(1,3)) +
                            geom_label_repel(data = pan8_cls, #change depending on Seurat used
                                 mapping = aes(x = U1,
                                               y=U2, 
                                               label = seurat_clusters), 
                                 size = 5, 
                                 min.segment.length = Inf, 
                                 show.legend = FALSE) +
                            guides(color = guide_legend(override.aes = list(alpha=1, 
                                                                          size=5, 
                                                                          shape=19)), 
                                   size = guide_legend()) +
                            scale_color_manual(values = custom_colors("clusters"),
                                               limits = force) #drops unused clusters in legend

p_cd8_panonly_treatments <- pan8@meta.data %>% 
                              ggplot(aes(x = UMAP_1, 
                                       y = UMAP_2, 
                                       color = seurat_clusters,
                                       size = pan_clonotype_count)) + 
                              geom_point(data = pan8@meta.data %>%
                                           dplyr::select(-treatment), 
                                           shape = 16,
                                           color = "grey90") +
                              geom_point(shape = 16, 
                                         alpha = 0.6) +
                              theme_minimal() +
                              theme(axis.title = element_blank(),
                                    axis.text = element_blank(),
                                    text = element_text(size = 15),
                                    panel.grid = element_blank(),
                                    legend.position = "none") +
                              labs(x = 'UMAP 1', 
                                   y = 'UMAP 2', 
                                   color = "Seurat clusters",
                                   title = "CD8+ T cells in the pancreas",
                                   size = "Size of clonotype\n(pancreas only)") +
                              scale_alpha_continuous(range = c(0.4,1)) +
                              scale_size_continuous(range=c(1,3)) +
                              facet_wrap(~treatment) +
                              scale_color_manual(values = custom_colors("clusters"),
                                                 limits = force) #drops unused clusters in legend

grid.arrange(p_cd8_panonly_cluster, p_cd8_panonly_treatments, widths = c(3,6))
```

Next, a stacked bar plot of the treatment group distribution within each seurat cluster:
```{r stacked bar chart of treatment groups per seurat cluster}
p_treatment_per_seurat_cluster <- pan8@meta.data %>% 
    ggplot(aes(x = factor(seurat_clusters, 
                        levels = rev(levels(pan8$seurat_clusters))), 
                        fill = treatment)) + 
    geom_bar(stat = "count", 
             position = "fill") + 
    scale_x_discrete(expand = c(0,0)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1), 
                       expand=c(0,0)) +
    theme_minimal() +
    theme(panel.grid = element_blank(),
          text = element_text(size = 15),
          legend.position = "right") +
    labs(title = "CD8+ T cells in the pancreas", 
         y = "Percent of Seurat Cluster", 
         x = NULL,
         color = "Tissue") +
    scale_fill_manual(values = custom_colors("treatments")) +
    coord_flip()
p_treatment_per_seurat_cluster
```

Bar chart showing the difference in the seurat clusters within each treatment group by each mouse. This is also shown by Fisher's exact test.
```{r bar chart of percent seurat clusters in each treatment group}
# p_seurat_cluster_per_treatment <- pan8@meta.data %>% 
#                                     group_by(treatment, mouse_id, seurat_clusters) %>%
#                                     tally() %>%
#                                     group_by(treatment,mouse_id) %>%
#                                     mutate(prcnt = n*100/sum(n)) %>%
#                                     ggplot(aes(x = factor(seurat_clusters, 
#                                                           levels = rev(levels(pan8$seurat_clusters))), 
#                                                fill = treatment,
#                                                y = prcnt)) + 
#                                     geom_violin(position = position_dodge(0.9), 
#                                                 scale = "width", 
#                                                 adjust = 1) + 
#                                     facet_wrap(~ seurat_clusters, 
#                                                nrow = 1, 
#                                                scales = "free_x") + 
#                                     scale_x_discrete(expand = c(0,0)) +
#                                     scale_y_continuous(expand=c(0,0)) +
#                                     theme_minimal() +
#                                     theme(panel.grid = element_blank(),
#                                           text = element_text(size = 15),
#                                           legend.position = "right",
#                                           panel.spacing = unit(0, "lines"),
#                                           panel.border = element_rect(fill = NA, color = "black", size = 1)) +
#                                     labs(title = "CD8+ T cells in the pancreas", 
#                                          y = "Percent of Seurat Cluster", 
#                                          x = NULL,
#                                          color = "Tissue") +
#                                     scale_fill_manual(values = custom_colors("treatments"))
# p_seurat_cluster_per_treatment
```

A volcano plot shows the DE genes between the PD1 and Spt treatment groups:
```{r volcano plot of pancreas-only differences PD1 and Spt}
cutoff <- 0.25 #set a cutoff for DE gene expression
ident1 <- "PD1"
ident2 <- "Spt"

Idents(pan8) <- pan8$treatment
volcanodata <- FindMarkers(pan8, #define the table of DE genes
                       ident.1 = ident1,
                       ident.2 = ident2, 
                       verbose = F,
                       only.pos = F, 
                       logfc.threshold = cutoff, 
                       assay = "RNA")
volcanodata$gene <- rownames(volcanodata) #rownames are genes
corner(volcanodata)

#assess significance
volcanodata$DE <- 'None' #set all significance to none
volcanodata[which(volcanodata$p_val_adj < 0.05 & volcanodata$avg_log2FC >= cutoff),"DE"] <- ident1 #positive avg_log2FC are associated with ident.1
volcanodata[which(volcanodata$p_val_adj < 0.05 & volcanodata$avg_log2FC <= -cutoff),"DE"] <- ident2 #negative avg_log2FC are associated with ident.2

#label individual genes
delabels <- volcanodata %>% filter(DE != 'None') %>% group_by(DE) #only label significant genes
delabels <- delabels %>% filter(gene %in% goi) #filter based on genes of interest

volcanodata %>%
    ggplot(aes(x = avg_log2FC,
               y = -log10(p_val_adj))) + 
        geom_hline(yintercept = -log10(0.05),
                   size = 0.5,
                   color = "grey90",
                   alpha = 0.5) + 
        geom_vline(xintercept = c(-cutoff, cutoff),
                   size = 0.5,
                   color = "grey90",
                   alpha = 0.5) + 
        geom_point(aes(color = DE),
                   size = 3,
                   alpha = 0.5,
                   shape = 16) +
        scale_color_manual(values = c("None" = "grey70",
                                      custom_colors("treatments"))) +
        scale_y_continuous(limits = c(0,125)) +
        theme_minimal() +
        theme(panel.grid = element_blank(),
              text = element_text(size = 15),
              axis.text = element_text(), 
              axis.title = element_text()) +
        labs(x = paste("Average Log2 Fold Change (",ident1,"/",ident2,")"),
             y = expression(-log[10]~(Adjusted~p~Value)),
             title="CD8+ T cells in pancreas of Spt") +
        geom_label_repel(data = delabels,
                         aes(x = avg_log2FC,
                             y = -log10(p_val_adj),
                             label = gene,
                             color = DE),
                         label.size = NA,
                         label.padding = 0.2, 
                         na.rm = TRUE,
                         fill = alpha(c("white"), 0.8), 
                         size=7, 
                         max.overlaps = 50)
```

Show the top 50 clonotypes as a percentage of TCRs detected as a cumulative sum. The clone sizes for the different treatment groups are also plotted. The diversity analysis is done by calculating the Simpson index without replacement, as it is less affected by differences in the absolute number of cells:
```{r top 50 clonotypes and clone size}
## TOP 50 CLONOTYPES
cumulative_sums <- pan8@meta.data %>% filter(!is.na(clonotype_id)) %>% 
                      group_by(treatment, clonotype_id, pan_clonotype_count) %>% 
                      tally() %>% 
                      group_by(treatment) %>% 
                      mutate(prcnt = n*100/sum(n)) %>% 
                      group_by(treatment) %>% 
                      arrange(treatment, desc(prcnt)) %>% 
                      mutate(cs = cumsum(prcnt))

cumulative_sums$clonotype_id <- factor(cumulative_sums$clonotype_id, 
                                       levels = cumulative_sums$clonotype_id)

p_top50_clonos <- cumulative_sums %>% 
                    group_by(treatment) %>% 
                    slice_head(n = 50) %>% 
                    ggplot(aes(x = clonotype_id, 
                               y = cs, 
                               color = treatment)) + 
                    geom_point() + 
                    geom_segment(aes(xend = clonotype_id, 
                                     yend = cs, 
                                     x = clonotype_id, 
                                     y = 0, 
                                     color = treatment)) + 
                    scale_color_manual(values = custom_colors("treatments")) +
                    scale_y_continuous(expand = c(0, 0), 
                                       limits = c(0, 80)) +
                    theme_minimal() +
                    theme(panel.grid = element_blank(),
                          panel.spacing.x = unit(0, "lines"),
                          text = element_text(size = 15),
                          axis.text.x = element_blank(),
                          panel.border = element_rect(colour = "black", 
                                                      fill = NA, 
                                                      size = 1),
                          legend.position = "none") +
                    labs(x = "Top 50 clonotypes", 
                         y = "Cumulative % CD8 TCRs",
                         title = "CD8+ T cells in pancreas") +
                    facet_wrap(~ treatment, scales = "free_x")
p_top50_clonos

## BOXPLOT OF CLONE SIZE
stat.test <- data.frame(pan8@meta.data %>%
                          dplyr::select(treatment, clonotype_id, pan_clonotype_count) %>%
                          unique() %>%
                          wilcox_test(pan_clonotype_count ~ treatment) %>% 
                          add_xy_position() %>%
                          mutate(y.position = log10(y.position + 1)))

pan8@meta.data %>%
    dplyr::select(treatment, clonotype_id, pan_clonotype_count) %>%
    unique() %>%
    ggplot() + 
    geom_boxplot(aes(x = treatment, 
               y = log10(pan_clonotype_count+1), 
               fill = treatment,
               color = treatment),
               outlier.shape = NA, 
               alpha = 0.5) +
    geom_point(aes(x = treatment, 
               y = log10(pan_clonotype_count+1), 
               fill = treatment,
               color = treatment),
               position = position_jitter(0.2),
               alpha = 0.1) +
    scale_fill_manual(values = custom_colors("treatments")) +
    scale_color_manual(values = custom_colors("treatments")) +
    theme_minimal() +
  scale_y_continuous(limits = c(0.3, 2.9)) +
    theme(panel.grid = element_blank(),
          axis.title.x = element_blank(),
          text = element_text(size = 15),
          panel.border = element_rect(colour = "black", fill=NA, size=1)) +
    labs(y = expression(log[10](clone~size~+1)),
         color = "Treatment",
         fill = "Treatment",
         title = "CD8+ T cells in pancreas\n(unique clonotypes)") +
    add_pvalue(stat.test,
               label = "p.adj.signif",
               step.increase = 0.05,
               tip.length = 0,
               label.size = 7) 

#DIVERSITY ANALYSIS
tcrtypecounts <- rbind(data.frame(pan8@meta.data %>% dplyr::select(treatment, mouse_id, clonotype_id, clonotype_count)))
corner(tcrtypecounts)

tcrtypecounts <- tcrtypecounts %>% filter(!is.na(clonotype_id)) %>% cast(treatment + mouse_id  ~ clonotype_id)
corner(tcrtypecounts)

totalN <- tcrtypecounts %>% select(-c(treatment, mouse_id)) %>% rowSums() # Total N by row sums
df <- sapply(tcrtypecounts %>% select(-c(treatment, mouse_id)), function(x) (x - 1)*x) # Calculate x(x-1) for each value in df
corner(df)
numerator <- df %>% rowSums() # Calculate numerator of equation (x in above)

# Calculate simpson index
diversities <- data.frame(tcrtypecounts %>% 
                            dplyr::select(c(treatment, mouse_id)), 
                          celltype = "CD8", 
                          simpson = totalN*(totalN - 1) / numerator)

diversities$treatment <- factor(diversities$treatment, levels = c("IgG", "Spt", "PD1")) #factor the pathologies

stat.test <- data.frame(diversities %>% 
                            mutate(simpson = ifelse(simpson == Inf, NA, simpson)) %>% 
                            wilcox_test(simpson ~ treatment) %>% 
                            add_xy_position())

diversities %>%
    ggplot(aes(x = treatment, 
               y = simpson,
               fill = treatment, 
               color = treatment)) + 
    geom_violin(alpha = 0.5,
                adjust = 1,
                scale = "width",
                trim = F) + 
    geom_point(size = 5) +
    labs(y = "Simpson index") +
    theme_minimal() +
    theme(panel.grid = element_blank(),
          text = element_text(size = 15),
          panel.border = element_rect(colour = "black", 
                                      fill = NA, 
                                      size = 1),
          legend.position = "bottom") +
    labs(title = "CD8 T cells in pancreas") +
    scale_fill_manual(values = custom_colors("treatments")) +
    scale_color_manual(values = custom_colors("treatments")) +
    add_pvalue(stat.test,
               inherit.aes = F,
               label = "p.adj.signif",
                y.position = "y.position") 

```

There is an outstanding question in the field as to whether diabetogenic T cells become exhausted in the pancreas. To explore this, co-inhibitory receptor expression is examined, in addition to expression of an exhaustion-associated gene signature:
```{r}
p_cd8_panonly_coinh_umap <- pan8@meta.data %>% 
                        ggplot(aes(x = UMAP_1, 
                                   y = UMAP_2, 
                                   color = coinh_count)) + 
                        geom_point(shape = 16, 
                                   alpha = 0.6) +
                        theme_minimal() +
                        theme(axis.title = element_blank(),
                              axis.text = element_blank(),
                              text = element_text(size = 15),
                              legend.position = c(0.85,0.85),
                              panel.grid = element_blank()) +
                        labs(x = 'UMAP 1', 
                             y = 'UMAP 2', 
                             color = "Co-inhibitory\nreceptors\nexpressed",
                             title = "CD8+ T cells in the pancreas") +
                        guides(color = guide_legend(override.aes = list(alpha=1, 
                                                                        size=5, 
                                                                        shape=19)), 
                               size = guide_legend()) +
                        scale_color_viridis(option = "E", direction = -1)

p_cd8_panonly_coinh_violin <- pan8@meta.data %>% 
                                ggplot(aes(x = treatment, 
                                         y = coinh_count, 
                                         fill = treatment,
                                         color = treatment)) + 
                                  geom_violin(adjust = 2, 
                                              scale = "width", 
                                              trim = F, 
                                              alpha = 0.5, 
                                              draw_quantiles = c(0.25,0.5,0.75), 
                                              lwd = 1) +
                                  scale_fill_manual(values = custom_colors("treatments")) +
                                  scale_color_manual(values = custom_colors("treatments")) +
                                  theme_minimal() +
                                  theme(panel.grid = element_blank(),
                                        axis.title.x = element_blank(),
                                        text = element_text(size = 15),
                                        panel.border = element_rect(colour = "black", fill=NA, size=1)) +
                                  labs(y = "Number of co-inhibitory receptors expressed",
                                       color = "Treatment",
                                       fill = "Treatment",
                                       title = "CD8+ T cells in pancreas")

# TERMINALLY EXHAUSTED MODULE SCORE ALL CELLS
p_vals <- pan8@meta.data %>% 
            wilcox_test(Miller2019_terminally_exh1 ~ treatment) %>%
            add_xy_position() %>%
            filter(p.adj <= 0.05)
pan8@meta.data %>% 
  ggplot(aes(x = treatment,
             y = Miller2019_terminally_exh1,
             color = treatment,
             fill = treatment)) + 
  geom_violin(adjust=2, 
              alpha=0.5, 
              draw_quantiles = c(0.25,0.5,0.75), 
              lwd=1) + 
  scale_color_manual(values = custom_colors("treatments")) + 
  scale_fill_manual(values = custom_colors("treatments")) +  
  theme_minimal() + 
  theme(text = element_text(size=15),
        panel.border = element_rect(fill = NA,
                                    color = "black"),
        axis.title.x = element_blank(),
        panel.grid = element_blank()) +
  labs(title = "Pancreas CD8+ T cells", 
       y = "Miller2019 terminally\nexhausted score",
       color = "Treatment",
       fill = "Treatment") + 
  add_pvalue(p_vals,
               inherit.aes = F,
               label.size = 8,
               step.increase = 0.05)

# TERMINALLY EXHAUSTED MODULE SCORE BY MOUSE
pan8@meta.data %>% 
    group_by(treatment,mouse_id) %>% 
    summarize(avgs = mean(Miller2019_terminally_exh1)) %>% 
    ggplot(aes(x = treatment,
               y = avgs,
               color = treatment,
               fill = treatment)) + 
    geom_violin(adjust = 1,
                trim = F,
                alpha = 0.5, 
                lwd = 1) + 
    geom_point(size = 5, shape = 21, stroke = 2, alpha = 0.2) +
    scale_color_manual(values = custom_colors("treatments")) + 
    scale_fill_manual(values = custom_colors("treatments")) +  
    theme_minimal() + 
    theme(text = element_text(size=15),
          panel.border = element_rect(fill = NA,
                                      color = "black"),
          axis.title.x = element_blank(),
          panel.grid = element_blank()) +
    labs(title = "Pancreas CD8+ T cells", 
         y = "Miller2019 terminally\nexhausted score",
         color = "Treatment",
         fill = "Treatment")
```

A Chi squared and Fisher's exact test will be used to measure the differences in the proportions of each cluster. This is done by looping through clusters and treatment groups to build a list of contingency tables:
```{r Fisher's exact test of proportions of clusters}
contingency_list <- list() #contingency tables as a list
df <- data.frame() #list of results from stats tests

#create a table containing raw count data
tbl <- pan8@meta.data %>% 
          group_by(treatment, seurat_clusters) %>% 
          tally() %>% 
          cast(seurat_clusters ~ treatment, value = "n")
tbl$seurat_clusters <- factor(tbl$seurat_clusters)
head(tbl)

#loop through treatments and clusters to build list of contingency tables
for (current_treatment in levels(pan8$treatment)) {
  for (current_cluster in levels(tbl$seurat_clusters)) {
    contingency_list[[current_treatment]][[current_cluster]] <- data.frame(treatment = c(pan8@meta.data %>% 
                               filter(seurat_clusters == current_cluster & treatment == current_treatment) %>% count() %>% pull(n),
               pan8@meta.data %>% filter(seurat_clusters == current_cluster & treatment != current_treatment) %>% count() %>% pull(n)),
               not_treatment = c(pan8@meta.data %>% filter(seurat_clusters != current_cluster & treatment == current_treatment) %>% count() %>% pull(n),
               pan8@meta.data %>% filter(seurat_clusters != current_cluster & treatment != current_treatment) %>% count() %>% pull(n)), 
               row.names = c("cluster","not_cluster"))
    df <- rbind(df, 
                data.frame(treatment = current_treatment, 
                           cluster = current_cluster, 
                           fisher.p.value = fisher.test(contingency_list[[current_treatment]][[current_cluster]])$p.value,
                           chisq.p.value = chisq.test(contingency_list[[current_treatment]][[current_cluster]])$p.value))
  }
}

#p values will be adjusted using "BH"
df$fisher.p.adj <- p.adjust(df$fisher.p.value, method = "BH")
df$chisq.p.adj <- p.adjust(df$chisq.p.value, method = "BH")

#factor some variables for subsequent plotting
df$treatment <- factor(df$treatment, levels = c("IgG","Spt","PD1"))
df$cluster <- factor(df$cluster, levels = levels(pan8$seurat_clusters))

p_fisher <- df %>% #plot of Fisher's Exact test results
  ggplot(aes(x = treatment, 
             y = cluster, 
             fill = -log10(fisher.p.adj))) + 
    geom_tile() +
    geom_text(aes(label = round(-log10(fisher.p.adj), 0))) + #round the adjusted p value
    scale_fill_distiller(direction = 1, 
                         palette = "Blues") +
    theme_minimal() +
    theme(text = element_text(size = 15),
          panel.grid = element_blank(),
          axis.title = element_blank(),
          legend.position = "top") +
    guides(fill = guide_colorbar(ticks.colour = NA)) +
    labs(title = "CD8+ T cells in pancreas\n(Fisher's Exact test)", 
         fill = expression(-~log[10]~(p~adjusted)))

p_chisq <- df %>% #plot of Chi-squared results
  ggplot(aes(x = treatment, 
             y = cluster, 
             fill = -log10(chisq.p.adj))) + 
    geom_tile() +
    geom_text(aes(label = round(-log10(chisq.p.adj), 0))) +
    scale_fill_distiller(direction = 1, 
                         palette = "Blues") +
    theme_minimal() +
    theme(text = element_text(size=15),
          panel.grid = element_blank(),
          axis.title = element_blank(),
          legend.position = "top",
          axis.text.y = element_blank()) + 
    guides(fill = guide_colorbar(ticks.colour = NA)) +
    labs(title = "CD8+ T cells in pancreas\n(Chi-squared p values)", 
         fill = expression(-~log[10]~(p~adjusted)))

#melt the raw data to plot
tbl <- tbl %>% 
        reshape2::melt(id.vars = "seurat_clusters")
colnames(tbl) <- c("cluster", "treatment", "cell_num") #meaningful column names

#plot the raw numbers
p_raw <- tbl %>% 
          ggplot(aes(x = treatment, 
                     y = cluster, 
                     fill = cell_num)) + 
            geom_tile() +
            geom_text(aes(label = cell_num)) +
            scale_fill_distiller(direction = 1, 
                                 palette = "Purples") +
            theme_minimal() +
            theme(text = element_text(size=15),
                  panel.grid = element_blank(),
                  axis.title = element_blank(),
                  legend.position = "top",) +
            guides(fill = guide_colorbar(ticks.colour = NA)) +
            labs(title = "Total CD8+ T cells in pancreas\n", 
                fill = "T cells in\ncluster")

grid.arrange(p_raw, p_fisher, p_chisq, ncol = 3, widths = c(1.5,1.5,1))
```

GSEA will be used to generate dot plots of enriched/interesting gene signatures that differ between CD8+ T cells across the treatment groups. For GSEA, we need the information of all genes, Seurat is just too slow if we test all 20,000 genes, so presto is used to perform a fast Wilcoxon rank sum test. The fgsea() function requires a list of gene sets to check, and a named vector of gene-level statistics, where the names should be the same as the gene names in the pathways list. A currated list of gene signatures will be examined.
```{r GSEA dot plots of enriched/interesting gene signatures}
all_genes <- wilcoxauc(pan8, "treatment") #fast Wilcoxon rank sum test between treatment groups

m_df <- rbind(msigdbr(species = "Mus musculus", category = "C7", subcategory = "IMMUNESIGDB")) #get database of gene signatures. Just using C7 IMMUNESIGDB
head(m_df)

fgsea_sets <- m_df %>% 
                split(x = .$gene_symbol, f = .$gs_name) #split the database into a list
corner(fgsea_sets)

#add custom gene signatures into fgsea
fgsea_sets$Miller2019_progenitor_exh <- read_excel("ref/signatures/Miller2019_NatureImmuno_SuppTable1.xlsx",sheet = "Progenitor Exh.")$gene
fgsea_sets$Miller2019_terminally_exh <- read_excel("ref/signatures/Miller2019_NatureImmuno_SuppTable1.xlsx",sheet = "Terminally Exh.")$gene
fgsea_sets$Miller2019_effector_like <- read_excel("ref/signatures/Miller2019_NatureImmuno_SuppTable1.xlsx",sheet = "Effector-like")$gene
fgsea_sets$beura_trm <- convertHumanGeneList(read_excel("ref/signatures/20200711_Beura_Trm_2017_Immunity.xlsx", sheet = "Beura_FRT_TRM_UP",col_names = F)$`...1`)
fgsea_sets$nonbystander <- read_excel("ref/signatures/20200718_Mognol_PNAS2017_bystanders.xlsx", sheet = "OT-1 UP (TS)",col_names = F)$`...1`
fgsea_sets$bystander <- read_excel("ref/signatures/20200718_Mognol_PNAS2017_bystanders.xlsx", sheet = "P14 UP (bystander)",col_names = F)$`...1`
fgsea_sets$cd8_activation_sarkar <- convertHumanGeneList(read.table("ref/signatures/cd8_activation_sarkar_up.txt")$V1)
fgsea_sets$cell_cycle <- convertHumanGeneList(read.table("ref/signatures/cell_cycle_all_strong.txt")$V1)
lag3 <- read_excel("ref/signatures/stephanie_de_genes.xlsx") #these DE genes are from FindAllMarkers, so they will be parsed out into separate signatures
  fgsea_sets$lag3exh1 <- lag3 %>% filter(cluster == 1) %>% filter(p_val_adj<0.05 & avg_logFC > 0) %>% pull(gene)
  fgsea_sets$lag3progexh2 <- lag3 %>% filter(cluster == 2) %>% filter(p_val_adj<0.05 & avg_logFC > 0) %>% pull(gene)
  fgsea_sets$lag3exh3 <- lag3 %>% filter(cluster == 3) %>% filter(p_val_adj<0.05 & avg_logFC > 0) %>% pull(gene)
  fgsea_sets$lag3act4 <- lag3 %>% filter(cluster == 4) %>% filter(p_val_adj<0.05 & avg_logFC > 0) %>% pull(gene)
  fgsea_sets$lag3act6 <- lag3 %>% filter(cluster == 6) %>% filter(p_val_adj<0.05 & avg_logFC > 0) %>% pull(gene)
  fgsea_sets$lag3eff7 <- lag3 %>% filter(cluster == 7) %>% filter(p_val_adj<0.05 & avg_logFC > 0) %>% pull(gene)
  fgsea_sets$lag3mem8 <- lag3 %>% filter(cluster == 8) %>% filter(p_val_adj<0.05 & avg_logFC > 0) %>% pull(gene)

genes <- list() #initialize a list of data.frames of features and auc for each treatment group
ranks <- list() #initialize a list of ranks for each treatment group
fgsea_results <- list() #initialize a list of gsea results
for (current_treatment in c("IgG","Spt","PD1")) {
  genes[[current_treatment]] <- all_genes %>%
                                  dplyr::filter(group == current_treatment) %>%
                                  arrange(desc(auc)) %>% 
                                  dplyr::select(feature, auc)
  write.table(genes[[current_treatment]], 
              row.names = F, 
              col.names = F, 
              sep = "\t", 
              quote = F,
              file = paste0("output/CD8_", current_treatment, ".rnk"))
  ranks[[current_treatment]] <- deframe(genes[[current_treatment]]) #converts two-column data frames to a named vector or list, using the first column as name and the second column as value
  fgsea_results[[current_treatment]] <- fgsea(fgsea_sets, #C7 gene signatures
                                              nperm = 1000,
                                              stats = ranks[[current_treatment]]) #use preranked genes
  fgsea_results[[current_treatment]] <- fgsea_results[[current_treatment]] %>% #tidy the table
                                          as_tibble() %>%
                                          arrange(desc(NES)) %>%
                                          dplyr::select(-leadingEdge, -ES, -nMoreExtreme) #remove unneccesary columns
}

#find the most interesting gene signatures
fgsea_results <- rbind(data.frame(fgsea_results[["PD1"]], treatment = "PD1"),
                  data.frame(fgsea_results[["Spt"]], treatment = "Spt"),
                  data.frame(fgsea_results[["IgG"]], treatment = "IgG"))

signatures_of_interest <- c("GSE9006_HEALTHY_VS_TYPE_1_DIABETES_PBMC_AT_DX_UP",
                            "GSE9006_HEALTHY_VS_TYPE_1_DIABETES_PBMC_4MONTH_POST_DX_UP",
                            "GSE9006_HEALTHY_VS_TYPE_1_DIABETES_PBMC_1MONTH_POST_DX_UP",
                            "GSE9006_TYPE_1_DIABETES_AT_DX_VS_1MONTH_POST_DX_PBMC_UP",
                            "GSE9006_TYPE_1_DIABETES_AT_DX_VS_4MONTH_POST_DX_PBMC_UP",
                            "GSE9650_EFFECTOR_VS_MEMORY_CD8_TCELL_UP",
                            "KAECH_DAY8_EFF_VS_MEMORY_CD8_TCELL_UP",
                            "KAECH_DAY8_EFF_VS_DAY15_EFF_CD8_TCELL_UP",
                            "GSE33425_CD161_INT_VS_NEG_CD8_TCELL_UP",
                            "GSE14699_DELETIONAL_TOLERANCE_VS_ACTIVATED_CD8_TCELL_UP",
                            "GSE41867_DAY6_EFFECTOR_VS_DAY30_MEMORY_CD8_TCELL_LCMV_ARMSTRONG_UP",
                            "GSE9650_EFFECTOR_VS_EXHAUSTED_CD8_TCELL_UP",
                            "GOLDRATH_EFF_VS_MEMORY_CD8_TCELL_UP",
                            "GSE30962_ACUTE_VS_CHRONIC_LCMV_PRIMARY_INF_CD8_TCELL_UP",
                            "GSE8678_IL7R_LOW_VS_HIGH_EFF_CD8_TCELL_UP",
                            "GSE8678_IL7R_LOW_VS_HIGH_EFF_CD8_TCELL_DN",
                            "Miller2019_progenitor_exh1",
                            "Miller2019_terminally_exh",
                            "Miller2019_effector_like",
                            "beura_trm",
                            "nonbystander",
                            "bystander",
                            "cell_cycle",
                            "cd8_activation_sarkar",
                            "lag3exh1",
                            "lag3progexh2",
                            "lag3exh3",
                            "lag3mem8",
                            "lag3eff7",
                            "lag3act4",
                            "lag3act6")

fgsea_results %>% 
  filter(pathway %in% signatures_of_interest & padj < 0.05) %>% 
  ggplot(aes(y = pathway,
             x = treatment,
             color = NES, 
             size = -log10(padj))) + 
  geom_point(shape=16) + 
  scale_size_continuous(range = c(1,10)) + 
  scale_color_gsea() +
  theme_minimal() + 
  theme(axis.title = element_blank(),
        panel.grid = element_blank(),
        text = element_text(size = 15))

melt(fgsea_results, id.vars = c("pathway","treatment"), measure.vars = c("padj","NES")) -> test
dcast(test, pathway ~ treatment + variable) -> test
```

Plot a bar chart of NRP-v7+ T cells per cluster
```{r plot NRP-v7 staining}
p_tetramer_per_cluster <- seurat_integrated@meta.data %>%
                            filter(!grepl("Tcon|Treg", seurat_clusters)) %>%
                            ggplot(aes(x = factor(seurat_clusters, 
                                                levels = rev(levels(seurat_integrated$seurat_clusters))), #orders clusters properly
                                                fill = tetramer_specificity)) + 
                            geom_bar(stat = "count", 
                                     position = "fill") + 
                            scale_x_discrete(expand = c(0, 0)) +
                            scale_y_continuous(labels = scales::percent_format(accuracy = 1), 
                                               expand = c(0, 0)) +
                            theme_minimal() +
                            theme(panel.grid = element_blank(),
                                  text = element_text(size = 15),
                                  legend.position = "right") +
                            labs(title = "CD8 T cells", 
                                 y = "Percent of Seurat Cluster", 
                                 x = NULL,
                                 color = "Tissue") +
                            scale_fill_manual(values = custom_colors("tetramer")) +
                            coord_flip()
p_tetramer_per_cluster_pan <- pan8@meta.data %>%
                            ggplot(aes(x = factor(seurat_clusters, 
                                                levels = rev(levels(seurat_integrated$seurat_clusters))), #orders clusters properly
                                                fill = tetramer_specificity)) + 
                            geom_bar(stat = "count", 
                                     position = "fill") + 
                            scale_x_discrete(expand = c(0, 0)) +
                            scale_y_continuous(labels = scales::percent_format(accuracy = 1), 
                                               expand = c(0, 0)) +
                            theme_minimal() +
                            theme(panel.grid = element_blank(),
                                  text = element_text(size = 15),
                                  legend.position = "right") +
                            labs(title = "CD8 T cells in pancreas", 
                                 y = "Percent of Seurat Cluster", 
                                 x = NULL,
                                 color = "Tissue") +
                            scale_fill_manual(values = custom_colors("tetramer")) +
                            coord_flip()
grid.arrange(p_tetramer_per_cluster, p_tetramer_per_cluster_pan, ncol = 2)
```